Projection variability: Is the family of sentences really a family?

### 2024-02-05 writing/SuB-paper/projection-operators-SuB.tex
WiP SuB paper, sec 2 graphs

\usepackage{mathptmx}
\usepackage{parskip}
\usepackage[top=1.5in,left=1in,right=1in,bottom=1in]{geometry}
%\setlength{\parindent}{0pt}
%\setlength{\parskip}{\the\baselineskip}
\pagestyle{empty}

@@

\usepackage{natbib} 
\bibpunct{(}{)}{;}{a}{,}{,}

@@

\usepackage{titleps}
\usepackage{lipsum}
\newpagestyle{mystyle}{
\sethead[\thepage][\Author][]{}{\Title}{\thepage}
}
\pagestyle{mystyle}
\author{Hofmann \and de Marneffe  \and Tonhauser}
\title{Projection variability of clausal complements across different operators}
\makeatletter
\newcommand\Author{Hofmann, de Marneffe, and Tonhauser}
\let\Title\@title
\makeatother
%header with author and title

\pagenumbering{gobble} 
%suppresses page numbering in header

\input{lisa-extra}

@@

{\large \textbf{Projection variability of clausal complements across different operators}}\footnote{We would like to thank\ldots Taylor Mahler; Reviewers at SALT, SuB, XPrag; Audience at Syntax \& Semantics colloquium at University of Stuttgart}\\
Lisa HOFMANN --- \textit{University of Stuttgart}\\
Marie-Catherine DE MARNEFFE --- \textit{UC Louvain}\\
Judith TONHAUSER --- \textit{University of Stuttgart}\\

\textbf{Abstract.}
	We present experimental evidence that the projection of clausal complements \textbf{(i)} varies between entailment-canceling operators, \textbf{(ii)} that the effect of operator varies between clause-embedding predicates, and \textbf{(iii)} we extend the result from \citet{degen_are_2022}, that projection ratings in polar questions are gradient in ways that cannot be predicted by categorical lexical classes, to contexts with negation, the epistemic possibility modal \textit{perhaps}, and conditional antecedents.
	%
	The observed variability is not captured by existing theoretical accounts of projection
	(e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}).
	%
	Our results suggest that an analysis must consider interactions between predicates and operators and raise important questions for future research on projection.

\textbf{Keywords:} projection, commitment, discourse interpretation.

\section{Introduction}
	Language users may infer that a speaker who utters an attitude ascription, as in \ref{ex:family}, is committed to the content of the complement (CC, here: \emph{Julian dances salsa,})
	even when it occurs under an entailment-canceling operator, like negation \ref{ex:neg}, polar questions \ref{ex:q}, epistemic possibility modals \ref{ex:mod}, or conditionals \ref{ex:cond}, in which case we say that it \textit{projects} (\citealt{frege_uber_1892,strawson_referring_1950,kiparsky_fact_1970,karttunen_conventional_1979}, and many more).

	\ex. \label{ex:family}
		\a. \label{ex:neg}
			{\bf Negation:} \hfill
			\emph{\lq Cole didn't discover that Julian dances salsa.\rq}
		\b. \label{ex:q}
			{\bf Polar Question:} \hfill
			\emph{\lq Did Cole discover that Julian dances salsa?\rq}
		\c. \label{ex:mod}
			{\bf Modal:} \hfill
			\emph{\lq Perhaps Cole discovered that Julian dances salsa.\rq}
		\d. \label{ex:cond}
			{\bf Conditional:} \hfill
			\emph{\lq If Cole discovered that Julian dances salsa, Logan will be joyful.\rq}
		\z.
	\z.

	We investigate the question whether the projection of embedded propositional content across entailment-cancelling operators differs by operator, and whether by-operator differences vary between embedding predicates.
	%
	Textbook diagnostics for projection often assume that, if a content projects across one of the operators in \Last, it uniformly projects across the others (\citealt{chierchia_meaning_1990}). Accordingly, current research examines projection variability between these operators only rarely, and with conflicting findings.

	\citet{karttunen_observations_1971} proposed a distinction between English factive predicates (e.g., \textit{be annoyed, regret, reveal, know}), where the CC projects across all four operators, and semi-factives (e.g., \textit{discover, realize, see, notice, find out}) where it always projects from under negation, but not always from under polar questions, modals, or conditionals.

	\citet{smith_relationship_2014} investigated by-operator variation, comparing negation and conditionals, for various types of projective content. They found that the projective content of epithets (e.g. \textit{idiot}) and the CC of \textit{know} was more projective under negation than conditionals, whereas that of appositive relative clauses and the preparatory content of \textit{win} showed the opposite pattern, and that of clefts showed no difference. Their task involved asking participants how surprised they would be to learn the content under investigation after observing the utterance. Since other factors besides speaker commitment modulate discourse (un)expectedness \citep[see e.g.][]{zimmermann_grammatical_2011,tonnis_german_2021}, it is unclear that they (only) measured projection.

	\posscite{sieker_projective_2022} study on German clause-embedding predicates compared the projection of purported factives (\textit{wissen} `know', \textit{bereuen} `regret', \textit{enthüllen} `reveal') and semi-factives (\textit{bemerken} `notice', \textit{entdecken} `discover', \textit{herausfinden} `find out') across the four contexts in \Last. They replicated Smith and Hall's result that the CC of \textit{know} projects more from under negation than conditionals for German \textit{wissen} (`know'). However, this finding was part of an overall pattern of higher projection ratings with negation than other operators. They found no interaction with predicate type, and therefore no evidence for the distinction of factives vs. semi-factives. This is based on an analysis which aggregated the ratings for the four purported factive and semi-factive predicates, under each of the entailment-cancelling operators.

	Studies in \citet{djarv_cognitive_2018}, \citet{tonhauser_how_2018}, and \citet{degen_are_2022} observed by-predicate variation in polar questions, where Karttunen's classification predicts more projection for factives than semi-factives.
	%
	\citet{djarv_cognitive_2018}, assessing acceptability of affirming the main clause while denying the CC, did find higher ratings for \emph{be happy} and \emph{appreciate} (assumed to be factive) and \emph{be aware} than \emph{realize} (assumed to be semi-factive). But here, it is also not obvious how exactly this task relates to projection.
	%
	\citet{tonhauser_how_2018} and \citet{degen_are_2022} measured projection of the CC of a broad range of attitude predicates more directly, collecting ratings about speaker certainty about the CC. The observed differences between individual clause-embedding predicates do not align with theoretically assumed lexical classes. While Degen \& Tonhauser demonstrate this for the distinction between purported factive and non-factive predicates, the results also did not match the expectations from Karttunen's factive vs. semi-factive classification.
	%
	Using speaker-certainty ratings, \citet{tonhauser_how_2018} found, e.g., that the CC of semi-factive \emph{realize} was as projective as that of factive \emph{be annoyed} and more than that of semi-factive \emph{discover}. \citet{degen_are_2022} observed that the CC of factive \textit{reveal} was less projective than that of semi-factive \textit{discover}, which, in turn, was less projective than that of factive \textit{know}. These findings suggest that individually examining projection ratings for various predicates, rather than pre-grouping them based on theoretical assumptions, can provide more nuanced insights into by-predicate variation in operator effects on projection.

	While \citet{smith_relationship_2014} found by-predicate variation in the effect of the operator on projection, \citet{sieker_projective_2022} did not.
	%
	These divergent results raise the question if they are due to cross-linguistic variation, task differences, or artifacts of the analysis. To address this, we conducted a series of experiments designed to assess projection across the four entailment-canceling operators in \ref{ex:family}. We used the same projection measure as \citet{sieker_projective_2022} (the `certain that' diagnostic; see e.g., \citealp{tonhauser_how_2018,djarv_prosodic_2017,mahler_social_2020}) and applied it to the CC of 20 English clause-embedding predicates, including factive (\emph{be annoyed, know, reveal}) and semi-factive predicates (\emph{discover, see}), and 15 non-factive predicates, given recent findings that their complements may also project, albeit to varying degrees (\citealt{degen_are_2022}).

\section{Assessing projection across entailment-canceling operators}

	To assess the effect of entailment-cancelling operators and clause-embedding predicates on the projection of the embedded content, we collected projectivity judgments for contents embedded under 20 clause-embedding predicates in four sets of experiments. The predicates were embedded under negation in Exps.~1, under polar questions (Exps.~2), under the epistemic possibility modal {\em perhaps} (Exps.~3), and in conditional antecedents (Exps.~4). (Each set of experiments contained three experiments differing in an at-issueness measure used in a separate block. We focus on the projection ratings here.)

		In each experiment, participants read utterances like those in \ref{ex:family} and judged whether the speaker (who was named) was certain of the CC (e.g.: Is [the speaker] certain that Julian dances salsa?).

		\begin{itemize}
			\item say more about task
		\end{itemize}

		Based on Karttunen's generalization, we would expect the CC of semi-factive predicates to receive high projection ratings under negation and lower ratings under the other operators, while the CC of factive predicates would consistently receive relatively high projection ratings throughout.

	\subsection{Experimental method}

		\subsubsection{Participants}
			We analyze the data from 2,682 self-reported native speakers of American English recruited on Prolific or Amazon's MT platform. (roughly 750 participants in each experiment)

		\subsubsection{Materials}
			Each participant saw all 20 attitude predicates (each paired with a unique content from a set of 20 contents) under one operator.

			\begin{itemize}
				\item 20 clause-embedding predicates that have shown projection variability in question contexts (Degen \& Tonhauser, 2022)
				\item Crossed w/ 20 CCs: 20 x 20 = 400 combinations
				\item One experiment per operator
				\item six main-clause controls for data exclusion
			\end{itemize}

@@

			On each trial, participants read an utterance, which was presented as an utterance of a speaker with a randomly chosen name, as well as the corresponding projection question, and then gave their response on a slider marked `no' (coded as 0) at one end and and `yes' (coded as 1) on the other. This is illustrated in Fig.~\ref{fig:trial}.

			Following \citet{tonhauser_how_2018}, we took a `yes' response to indicate that participants interpret the utterance in a way that the presented speaker (e.g., Christopher in Fig.~\ref{fig:trial}) is committed to the embedded content, which reflects projection. A `no' response was taken to indicate that the content does not project.

			\begin{figure}[ht]
				\centering
				\includegraphics[width = .7\linewidth]{task-1n-proj.eps}
				\caption{A sample trial in Exp.~1. In the corresponding trials in Exps.~2-4, participants were presented with an utterance involving a different entailment-cancelling operator.}
				\label{fig:trial}
			\end{figure}

	\subsection{Results and discussion}
		Our analysis reveals two key results: \textbf{(i)} There is projection variability by operator and \textbf{(ii)} there is by-predicate variation in the effect of operator on projection.

		\subsubsection{By-operator variation}

			We find an effect of operator when aggregating across predicates. Mean projection ratings were higher under question-embeddings than under negation and modals, but lower than in conditional antecedents. This is shown in Fig.~\ref{fig:op-ratings}, and the generalizations are supported by Model \#1 reported in Table~\ref{t:op-model}.

			\begin{figure}[ht]
				\centering
				\includegraphics[width = .35\linewidth]{projective-op}
				\caption{Mean certainty ratings by operator with $95\%$ bootstrapped confidence intervals.}
				\label{fig:op-ratings}
			\end{figure}

@@

					\caption{Excerpt from the model output of a linear mixed effects model with fixed effects of operator; random effects: participant and item intercepts, fit with \texttt{lme4, lmertest} in \texttt{R}.\label{t:op-model}}
				\end{table}

			This result differs from that of \cite{sieker_projective_2022}, who observed that the CC of German attitude predicates projects more from under negation than the other operators.

			\begin{itemize}
				\item but like them, we observe small differences 
				\item overall, they have much higher projection ratings
				\item but they chose a different set of predicates, we can understand these different results when taking into account by-predicate variation in the effect of operator
			\end{itemize}

		\subsubsection{By-predicate variation in the effect of operator}

			We observe differences across the predicates in by-operator variation. This is illustrated in Figure~\ref{fig:op-pred-ratings}, which shows mean projection ratings for the 20 attitude predicates by embedding operator; predicates are ordered by their mean rating across all operators (\emph{be annoyed} has the highest overall mean).
			%
			For instance, whereas the CC of \emph{be annoyed} projects more from under negation and questions than conditionals and modals, and the CC of \emph{discover} projects less from negation than conditionals and questions, and more from under negation than modals. The CC of \emph{know} projects less from under negation than questions, but more from under negation than modals, while the difference between negation and conditionals is not significant.
			%
			These generalizations are supported by models\ \# 2--4 in Table\ \ref{t:models}, which also each have at least $34$ highly significant interaction terms (out of $57$ possible interactions of operator and predicate).

>>>>

\usepackage{mathptmx}
\usepackage{parskip}
\usepackage[top=1.5in,left=1in,right=1in,bottom=1in]{geometry}
%\setlength{\parindent}{0pt}
%\setlength{\parskip}{\the\baselineskip}

% JT: I've commented this out so that we can see page numbers while writin
%\pagestyle{empty}

@@

\usepackage{natbib} 
%\bibpunct{(}{)}{;}{a}{,}{,}
\bibpunct{(}{)}{,}{a}{,}{,}

@@

\usepackage{titleps}
\usepackage{lipsum}
\newpagestyle{mystyle}{
\sethead[\thepage][\Author][]{}{\Title}{\thepage}
}
\pagestyle{mystyle}
\author{Hofmann \and de Marneffe  \and Tonhauser}
%\title{Projection variability of clausal complements across different operators}
\title{Projection variability: Is the family of sentences really a family?}
\makeatletter
\newcommand\Author{Hofmann, de Marneffe, and Tonhauser}
\let\Title\@title
\makeatother
%header with author and title

% JT commented out whle editing
%\pagenumbering{gobble} 
%suppresses page numbering in header

%% other packages I added (LH)
% misc formatting
	\usepackage{booktabs}
    \usepackage[export]{adjustbox}
    \usepackage{rotating}

%% bibliography
	\newcommand{\posscite}[1]{\citeauthor{#1}'s (\citeyear{#1})}

 \newcommand{\poscite}[1]{\citeauthor{#1}'s \citeyear{#1}}

% drawing
	\usepackage{graphicx}
	\usepackage{xcolor}

@@

%{\large \textbf{Projection variability of clausal complements across different operators}}
{\large \textbf{Projection variability: Is the family of sentences really a family?}}\footnote{We thank Taylor Mahler for assistance in collecting the data presented here as well as valuable comments. We gratefully acknowledge the National Science foundation grant \#1452674 (to Marie-Catherine de Marneffe, Craige Roberts, and Judith Tonhauser), which provided financial support for this project.}\\
Lisa HOFMANN --- \textit{University of Stuttgart}\\
Marie-Catherine DE MARNEFFE --- \textit{UCLouvain}\\
Judith TONHAUSER --- \textit{University of Stuttgart}\\

\textbf{Abstract.} We present experimental evidence that the projection of the content of clausal complements \textbf{(i)} varies across entailment-canceling operators, \textbf{(ii)} that the effect of operator varies between clause-embedding predicates, and \textbf{(iii)} we extend a result of \citealt{degen_are_2022}, that projection ratings in polar questions do not categorically distinguish factive from nonfactive predicates, to cases with negation, the epistemic possibility modal \textit{perhaps}, and conditional antecedents. The observed variability is not captured by existing theoretical accounts of projection (e.g., \citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}). Our results suggest that an analysis must consider interactions between predicates and operators and raise important questions for future research on projection.

\textbf{Keywords:} Projection variability, entailment-canceling operators, (non)factive predicates.

\section{Introduction} \label{sec:intro}

    Textbooks typically present the family-of-sentences diagnostic for projection without addressing by-operator variation (e.g., \citealt{chierchia_meaning_1990}). For instance, the clausal complement of \emph{discover} is diagnosed as as projective content, if its content (here: that Julian dances salsa) follows not just from \ref{ex:simple}, but also from its variants in \ref{ex:family}, where \ref{ex:simple} has been embedded under an entailment-canceling operator, such as negation \ref{ex:neg}, polar questions \ref{ex:q}, epistemic possibility modals \ref{ex:mod}, or conditional antecedents \ref{ex:cond}.

    % In formal semantics/pragmatics, the so-called `family of sentences' diagnostic is the standard way of diagnosing whether a content is projective (e.g., \citealt{chierchia_meaning_1990}). For instance, to diagnose the content of the clausal complement of \emph{discover} as projective content, a researcher would typically observe that the content that Julian dances salsa follows not just from \ref{ex:simple}, but also from the family-of-sentences variants of \ref{ex:simple} in \ref{ex:family}, where \ref{ex:simple} has been embedded under an entailment-canceling operator, such as negation \ref{ex:neg}, polar questions \ref{ex:q}, epistemic possibility modals \ref{ex:mod}, or conditional antecedents \ref{ex:cond}.

    % (e.g., \citealt{frege_uber_1892,strawson_referring_1950,kiparsky_fact_1970,karttunen_conventional_1979}).

    \ex.\label{ex:simple} Cole discovered that Julian dances salsa.

	\ex. \label{ex:family}
		\a. \label{ex:neg}
			{\bf Negation:} \hfill
			\emph{Cole didn't discover that Julian dances salsa.}
		\b. \label{ex:q}
			{\bf Polar Question:} \hfill
			\emph{Did Cole discover that Julian dances salsa?}
		\c. \label{ex:mod}
			{\bf Modal:} \hfill
			\emph{Perhaps Cole discovered that Julian dances salsa.}
		\d. \label{ex:cond}
			{\bf Conditional:} \hfill
			\emph{If Cole discovered that Julian dances salsa, Logan will be joyful.}
		\z.
	\z.

    However, prior research is suggestive of variation in the effect of different entailment-cancelling operators on projection. For instance, \citet{karttunen_observations_1971} proposed a distinction between English factive predicates (e.g., \textit{regret}) and semi-factives (e.g., \textit{discover}).
    Based on the examples in \Next, he suggested that the CC of true factives projects across all four operators, but that of semi-factives does not always project from under polar questions, modals, or conditionals.\footnote{See \citealt{beaver_have_2010} for naturally occurring examples in which the CCs of factive predicates do not project from under the four operators, thereby calling into question the distinction between factive and semi-factive predicates. For experimental research on the distinction see \citealt{djaerv-etal2016} and \citealt{djarv_cognitive_2018}. We return to this literature in section \ref{s:analysis}.}

    \ex. \citealt{karttunen_observations_1971}: (22, 24--26)
        \a. John didn't \{regret/discover\} that he had not told the truth.
        \b. Did you \{regret/discover\} that you had not told the truth?
        \b. If I \{regret/discover\} later that I have not told the truth, I will confess it to everyone.
        \b. It is possible that I \{regret/discover\} later that I have not told the truth.
        \z.
    \z.

    There have been two experimental investigations of by-operator projection variation. First, the experiment reported in \citet{smith_relationship_2014} investigated projection from under negation and antecents of conditionals for various types of English projective contents. They found that the expressive content of epithets (e.g., \textit{idiot}) and the CC of \textit{know} was more projective under negation than conditionals. In contrast, the content of appositive relative clauses and the preparatory content of \textit{win} showed the opposite pattern, and the existential presupposition of clefts showed no difference. %However, their task involved asking participants how surprised they would be to learn the content under investigation after observing the utterance. Since other factors besides speaker commitment modulate discourse (un)expectedness (see e.g., \citealt{zimmermann_grammatical_2011,tonnis_german_2021}), it is unclear that they (only) measured projection.

    Second, \citet{sieker_projective_2022}  compared the projection of German factives (\textit{wissen} `know', \textit{bereuen} `regret', \textit{ent\-hüllen} `reveal') and semi-factives (\textit{bemerken} `notice', \textit{entdecken} `discover', \textit{herausfinden} `find out') between the four operators in \ref{ex:family}. The results showed that the CCs of the predicates investigated projected more from under negation than the other three operators. contrary to what \citealt{karttunen_observations_1971} suggested, a comparison of factive and semifactive predicates did not reveal that the CCs of factive predicates project more from under polar questions, modals, or antecedents of conditionals than the CCs of nonfactive predicates.

    %They replicated \poscite{smith_relationship_2014} result that the CC of \textit{know} projects more from under negation than conditionals for German \textit{wissen} (`know'). However, this finding was part of an overall pattern of higher projection ratings with negation than other operators.
    %
    %Grouping ratings for the four factive and semi-factive predicates, respectively, they found no interaction of operator with predicate type, and thus no evidence for the factive/semi-factive distinction (iii), or an operator/predicate interaction (ii). However, their analysis did not explore the impact of operators on each predicate individually.

    This paper reports on the results of a set of experiments that were designed to compare projection from under the four entailment-canceling operators in \ref{ex:family}. Our experiments extend the empirical scope of prior research by investigating projection for a broader set of contents, namely the contents of the complements of the 20 English clause-embedding predicates in \ref{ex:predicates}:

    \ex. \label{ex:predicates}
    		\a. (Semi)factive predicates: {\em be annoyed, know, reveal, discover, see}
      \b. Nonfactive predicates: {\em acknowledge, admit, announce, confess, confirm, establish, hear, inform, prove, be right, demonstrate, pretend, say, suggest, think}

    As shown, the  20 predicates include both (semi)factive and nonfactive predicates. The inclusion of the latter set of predicates is motivated by the results of the empirical investigations in \citealt{degen_are_2022}, which suggested that the contents of complements of nonfactive predicates is also projective and that projection does not categorically distinguish factive and nonfactive predicates. As \poscite{degen_are_2022} result was established based on sentences where the predicates were embedded under polar questions, our experiments, which feature all four entailment-canceling operators, serve to replicate and extend this result of \citealt{degen_are_2022}.

    %    The factive/semi-factive distinction (iii) is further called into question by \citealt{beaver_have_2010}, who reports many naturally occurring examples of cognitive factive predicates, showing that their CC does not necessarily project (e.g., English \emph{be/become aware, know, discover, notice, find out, realize, accept the fact}; French \emph{trouver} \lq discover\rq).

    %Experimental findings from \citet{djarv_cognitive_2018}, \citet{tonhauser_how_2018}, and \citet{degen_are_2022} can be evaluated with respect to whether they support Karttunen's distinction (iii), with mixed results. These studies observed by-predicate variation in polar questions, where more projection is expected for factives than semi-factives.
	%
	%\citet{djarv_cognitive_2018}, assessing acceptability of affirming the main clause while denying the CC, did find higher ratings for English \emph{be happy} and \emph{appreciate} (assumed to be factive) and \emph{be aware} than \emph{realize} (assumed to be semi-factive). This could be interpreted as limited support for (iii). But here, it is also not obvious how exactly this task relates to projection.\footnote{Say why}

	%\citet{tonhauser_how_2018} and \citet{degen_are_2022} measured speaker commitments toward the CC of a broad range of English clause-embedding predicates more directly, by collecting judgments of how certain participants take a speaker to be of the contents under investigation.
	%
	%The observed differences between the various predicates do not align with theoretically assumed lexical classes. While \citet{degen_are_2022} demonstrate this for the distinction between purported factive and non-factive predicates, the results did not match the expectations from Karttunen's factive vs. semi-factive classification either.
	%
	%\citet{tonhauser_how_2018} found, e.g., that the CC of semi-factive \emph{realize} was as projective as that of factive \emph{be annoyed} and more than that of semi-factive \emph{discover}. \citet{degen_are_2022} observed that the CC of factive \textit{reveal} was less projective than that of semi-factive \textit{discover}, which, in turn, was less projective than that of factive \textit{know}. These findings also suggest that examining projection ratings for various predicates individually, rather than pre-grouping them based on theoretical assumptions, can provide more nuanced insights into by-predicate projection variation.

	%Both \citet{smith_relationship_2014} \citet{sieker_projective_2022} found that projection is affected differently by entailment-cancelling operators. However, while Smith \& Hall found variation in the effect of the operator on projection between various projective contents, Sieker \& Solstad did not.
	%
	%These divergent results raise the question if they are due to cross-linguistic variation, task differences, or artifacts of the analysis. Finally, much previous research calls into question the hypothesis that by-predicate differences in the effect of operator on projection pattern according to Kattunen's factive/semi-factive distinction (\citealt{beaver_have_2010,tonhauser_how_2018,sieker_projective_2022,degen_are_2022}).

  %Karttunen's distinction gives rise to three hypotheses about the projection associated with clause-embedding predicates, which we explore in this paper: \textbf{(i) by-operator variation:} entailment-cancelling operators differentially affect projection; \textbf{(ii) operator/predicate interaction:} the effect of operator differs between predicates; \textbf{(iii) factive/semi-factive distinction:} this difference can be addressed by characterizing lexical classes of (semi-)factive predicates.

    %To address the three hypotheses, we conducted a set of experiments designed to assess projection across the four entailment-canceling operators in \ref{ex:family} for various clause-embedding predicates. We used the same projection measure as \citet{sieker_projective_2022} (the `certain that' task; see e.g., \citealp{tonhauser_how_2018,djarv_prosodic_2017,mahler_social_2020}) and applied it to the CC of 20 English predicates, including factives and semi-factives, and 15 non-factive predicates, given recent findings that their complements may also project, albeit to varying degrees (\citealt{degen_are_2022}). Our findings support (i) by-operator variation, and an operator/predicate interaction (ii), but not Karttunen's factive/semi-factive distinction (iii).

\section{Experiments}

	To assess the effect of entailment-cancelling operator and clause-embedding predicate on projection, we collected projection judgments for the CCs of the 20 clause-embedding predicates in four sets of experiments.\footnote{The data and analysis scripts can be found in the following GitHub repository: \url{https://github.com/judith-tonhauser/CommitmentBankPlus}. {\bf LETS BE MORE SPECIFIC HERE ABOUT WHERE THINGS RELATED TO THIS PAPER ARE}} The predicates were embedded under negation in Exps.~1, under polar questions in Exps.~2, under the epistemic possibility modal {\em perhaps} in Exps.~3, and in conditional antecedents in Exps.~4. Each set of experiments consisted of three experiments that differed in the at-issueness measure used in a separate block. In this paper, we limit our attention to the projection ratings collected in the 12 experiments.

	Projection was measured in each experiment with the `certain that' diagnostic, which has been used to measure projection with both polar interrogative and declarative sentences (see, e.g., \citealt{tonhauser_prosodic_2016,djarv_prosodic_2017,stevens_rational_2017,lorson_influence_2018,tonhauser_how_2018,mahler_does_2019,mahler_social_2020,de_marneffe_commitmentbank_2019}).\footnote{For other diagnostics of projection see, e.g.\ \citealt{smith_projection_2011,xue_correlation_2011}, and \citealt{tonhauser_toward_2013}; and discussion in \citealt{tonhauser_how_2018}.} This diagnostic was also used in \citealt{sieker_projective_2022}. Under this diagnostic, participants are presented with utterances like those in \Next. They are asked to rate whether the (named) speaker is certain of the CC.

	\ex. \label{ex:certain-that}
		\a. Christopher: \emph{\lq Cole didn't discover that Julian dances salsa.\rq}
		\b. Christopher: \emph{\lq Did Cole discover that Julian dances salsa?\rq}
		\c. Christopher: \emph{\lq Perhaps Cole discovered that Julian dances salsa.\rq}
		\d. Christopher: \emph{\lq If Cole discovered that Julian dances salsa, Logan will be joyful.\rq}
		\z.
		Projection question: Is Christopher certain that Julian dances salsa?
	\z.

    We assume, following \citealt{tonhauser_how_2018} and \citealt{degen_are_2022}, that judgments of speaker certainty about some propositional content reflect speaker commitment to that content, that is, projection. If a participant interprets an utterance like \Last[a--d] in a way that the speaker (e.g., Christopher) is certain of the CC, it is assumed to project out of the entailment-canceling environment. If a participant does not take the speaker to be certain of the CC, this is taken to indicate that it does not project.

    %This setup allows us to test the three hypotheses outlined in Section \ref{sec:intro}. First, a main effect of operator on speaker certainty judgments for embedded contents would provide evidence for by-operator variation (i). Second, if there is a operator/predicate interaction (ii), we would expect the effect of operator to be different (in magnitude or direction) for different predicates. Finally, based on the factive/semi-factive distinction (iii), we would expect the CC of semi-factive predicates to receive high projection ratings under negation and lower ratings under the other operators, while the CC of factive predicates would consistently receive relatively high projection ratings throughout.

	\subsection{Methods}

		\subsubsection{Participants}
			We recruited 250-300 participants for each of the 12 experiments. Participants for Exp.~1q were recruited on Amazon's Mechanical Turk platform; the participants were required to have U.S.\ IP addresses and at least 99\% of previously approved HITs. Participants for the remaining experiments were recruited on Prolific; these participants were required to reside in the US, to be born in the US, to have English as their first language, and to have an approval rating of at least 99\%. See Supplement \ref{app:d-participants} for further information about the participants.

		\subsubsection{Materials}

            The target sentences consisted of 400 combinations of the 20 clause-embedding predicates in \ref{ex:predicates} with 20 embedded clauses (provided in Supplement \ref{app:a-clauses}). As mentioned above, the predicates were embedded under negation in Exps.~1, under polar questions in Exps.~2, under the epistemic possibility modal \emph{perhaps} in Exps.~3, and in conditional antecedents in Exps.~4, for a total of 400 target stimuli in each of the four sets of experiments.

			%We tested the 20 clause-embedding predicates in \Next, which have shown projection variability in polar questions (\citealt{degen_are_2022}). They comprise three factive predicates \Next[a], two semi-factives \Next[b], nine that have been characterized as optionally factive in \citealt{kiparsky_fact_1970} \Next[c], and six non-factives which include two veridical predicates \Next[d], and four non-veridical ones \Next[e].

	%		\ex. 20 clause-embedding predicates \hfill \citealt{degen_are_2022}
			%	\a. canonically factive:					\emph{be annoyed, know, reveal}
				%\b. semi-factive: 		\emph{discover, see}
%				\b. optionally factive: \newline \emph{acknowledge admit, announce, confess, confirm, establish, hear, inform, prove}
%				\b. Non-factive veridical:
%					\emph{be right, demonstrate}
%				\b. Non-factive non-veridical:
%				\emph{pretend, say, suggest, think}
%				\z.
%			\z.

To assess whether participants were attending to the task, each experiment included six control stimuli. For details on the six control stimuli see Supplement \ref{app:c-control}.

Each participant saw a random set of twenty-six stimuli: Each set contained one target stimulus for each of the 20 clause-embedding predicates (each with a unique complement clause) and the same six control stimuli.\footnote{Each participant saw their set of 26 stimuli twice, once in the projection block and once in the at-issueness block. Block order was randomized. As mentioned above, we focus on the projection ratings here.} Trial order was randomized.

  %which were main clauses with appositive relative clauses, which were also presented as utterances made by a named speaker. Here, the main clause content was targeted by the projection question. In the experiments with polar questions, the controls were expected to receive low certainty ratings, whereas in the experiments with declaratives, they were expected to receive high ratings.

  %Each participant saw all 20 predicates, randomly paired with a unique embedded clause from a set of 20 contents, and a subject from a set of 20 proper names. %The predicates associated with projective content were embedded under (i) negation, (ii) polar questions, (iii) the epistemic possibility modal \textit{perhaps}, and (iv) in conditional antecedents in the four different sets of experiments.

@@

			Participants were told to imagine that they are at a party and that, on walking into the kitchen, they overhear somebody say something to somebody else.
            %
            On each trial, participants read an utterance and responded to the `certain that'   question. They gave their response on a slider marked `no' (coded as 0) at one end and and `yes' (coded as 1) on the other. A sample trial from Exps.~1 is illustrated in Figure~\ref{fig:trial}. Following, e.g., \citet{tonhauser_how_2018}, higher ratings of speaker certainty could reflect one of two things:
            %
            On the one hand, higher ratings could reflect greater speaker commitment towards the CC, and therefore greater projection. This assumes that projection (and the relevant notion of speaker commitment) can itself be gradient.
            %
            Conversely, higher ratings could also reflect the probability that a participant associates with a projective interpretation for the utterance. Under this interpretation, commitment may be interpreted categorically, and the gradience is a result of the uncertainty about possible interpretations (see also discussion in \citealt{grove_factivity_2023}, which suggests the latter kind of interpretation).

            %  we take high ratings of speaker certainty to reflect speaker greater speaker commitment towards the CC, that is, greater projection.

			\begin{figure}[ht]
				\centering
				\fbox{\includegraphics[width = .7\linewidth]{task-1n-proj.eps}}
				\caption{A sample trial in Exps.~1. In the corresponding trials in Exps.~2-4, participants were presented with an utterance with a different entailment-cancelling operator.}
				\label{fig:trial}
			\end{figure}

             After completing the experiment, participants filled out a short optional demographic survey. To encourage truthful responses, participants were told that they would be paid no matter what answers they gave in the survey.

        \subsubsection{Data exclusion}

        Data were excluded based on self-declared non-native speaker status and other criteria given in the Supplements \ref{app:d-participants}. The data from 2,682 participants entered into the analysis.

            %We analyze the data from $2,682$ self-reported native speakers of American English, after excluding participants who failed one of the following two attention-checks: their mean rating on six control items is more than 2 sd different from the group mean, or they always roughly selected the same point on the scale for target stimuli (identified by low variance, and then manually inspecting their response patterns).

    \subsection{Results and discussion}

    We first address the question of whether there is by-operator variation (section \ref{s:by-op}) and then the question of whether there is by-predicate variation in the observed by-operator variation (section \ref{s:by-pred}).

		%Our data reveals three key results: \textbf{(i)} There is projection variability by operator; \textbf{(ii)} There is by-predicate variation in the effect of operator on projection; and \textbf{(iii)} We find further support (from entailment-cancelling operators beyond polar questions) for \posscite{degen_are_2022} result that by-predicate projection variation is gradient in a way that cannot be predicted by categorical lexical classes.

	\subsubsection{By-operator variation}\label{s:by-op}

        Fig.~\ref{fig:op-ratings} shows the mean certainty ratings by entailment-canceling operator, aggregating over the clause-embedding predicates. As shown, there is projection variability by operator: The CCs of the clause-embedding predicates were relatively more projective when embedded in the antecedent of a conditional than in a polar interrogative, where they were relative more projective than when they were embedded under negation or the epistemic modal {\em perhaps}.

			%When aggregating across predicates, we found an effect of \texttt{operator}. The distribution of projection ratings, as well as means and $95\%$ bootstrapped confidence intervals by operator are shown in Figure~\ref{fig:op-ratings}.

			\begin{figure}[ht]
				\centering
				\includegraphics[scale = .8]{operator-graph-1}
				\caption{Mean certainty ratings by operator. Error bars indicate $95\%$ bootstrapped confidence intervals. Violin plots indicate the kernel probability density of the individual participants’ ratings.}
				\label{fig:op-ratings}
			\end{figure}

			%Mean projection ratings were higher for CCs in question-embeddings than under negation and modals, but lower than in conditional antecedents.

   These generalizations are supported by Model \#1 reported in Table~\ref{t:op-model}.

@@

					\caption{Excerpt from the model output of a linear mixed effects model with fixed effects of operator; random effects: participant and item intercepts, fitted with \texttt{lme4, lmertest} in \texttt{R}.\label{t:op-model}}
				\end{table}

			Like \citet{sieker_projective_2022}, we observe a small main effect of entailment-cancelling operator on projection. However, in contrast to their result that projection is strongest from under negation, our results suggest that projection is strongest from antecedents of conditionals.
            We note, however, that the differences in mean certainty ratings between the four entailment-canceling operators are very small. This suggests that, when abstracting away from individual projective contents, projection from under the four entailment-canceling operators is very similar. In other words, when abstracting away from individual projective contents, the family of sentences really are a family.

		\subsubsection{By-predicate variation in the effect of entailment-canceling operator}\label{s:by-pred}
